Software architectures to integrate workflow engines in science gateways

### 2016-05-15 paper.tex
Major updates following comments by @SilviaOlabarriaga and @vkorkhov

\author[mcgill]{Reza Adalat}
\author[mcgill]{Natacha Beck}
\author[mcgill]{\\Samir Das}
\author[isi]{Rafael Ferreira da Silva}
\author[mcgill]{Najmeh Khalili-Mahani}
\author[criugm]{Pierre-Olivier Quirion}
\author[mcgill]{Pierre Rioux}
\author[amc]{S\'ilvia D. Olabarriaga}
\author[criugm]{Pierre Bellec}
\author[mcgill]{Alan C. Evans}

\address[mcgill]{McGill Centre for Integrative Neuroscience, Montreal Neurological Institute, McGill University, Canada.}
\address[creatis]{University of Lyon, CNRS, INSERM, CREATIS, Villeurbanne, France.}
\address[criugm]{Centre de Recherche de l'Institut de G\'eriatrie de Montr\'eal CRIUGM, Montr\'eal, QC, Canada.}
\address[isi]{University of Southern California, Information Sciences Institute, Marina del Rey, CA, USA.}
\address[amc]{Academic Medical Center of the University of Amsterdam, Department of Clinical Epidemiology, Biostatistics and Bioinformatics, Amsterdam, NL.}

\begin{abstract}
  Workflow engines are critical for the efficient and transparent
  exploitation of distributed infrastructures in the ecosystem of
  tools and services offered by science gateways.  We investigate six
  software architectures commonly used to integrate workflow engines
  into science gateways. In \emph{tight integration}, the workflow
  engine shares software components with the science gateway. In
  \emph{service invocation}, the engine is isolated and invoked
  through a specific software interface. In \emph{task encapsulation},
  the engine is wrapped as a computing task executed on the
  infrastructure. In the \emph{pool model}, the engine is bundled in
  an agent that connects to a central pool to fetch and execute
  workflows. In \emph{nested workflows}, the engine is integrated as a
  child process of another engine. In \emph{workflow import}, the
  engine is integrated through workflow language conversion. We
  describe and evaluate these architectures with metrics for
  assessment of integration complexity, robustness, extensibility,
  scalability and support for meta-workflows and fine-grained
  debugging. Tight integration and task encapsulation are the easiest
  to integrate and the most robust. Extensibility is equivalent in
  most architectures. The pool model is the most scalable one and
  meta-workflows are only available in nested workflows and workflow
  import. These results provide insights for science gateway
  architects and developers.
\end{abstract}

@@

This paper describes, provides examples and compares such
architectures, based on system-independent representations of their
main components and interactions. It is informed by our experince in
the development and sustained operation of the CBRAIN~\cite{SHER-14}
and VIP~\cite{GLAT-13} science gateways for medical image analysis
during the past 7 years, as well as by lessons learned from several
science gateway and workflow projects such as
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}}.  In this section
we provide background information and definitions of 
workflow engines, science gateways and infrastructures. In
Section~\ref{sec:architectures}, we describe six architectures within
a consistent framework that underlines the functional interactions
between their main software components. In
Section~\ref{sec:evaluation}, the different architectures are
evaluated with metrics measuring integration complexity, robustness,
extensibility, scalability and other specific features. We finally
compare the architectures and highlight recommendations for science
gateway architects and workflow engine developers.

@@

In the last decade, the e-Science community has developed workflow
systems to help application developers access distributed
infrastructures such as clusters, grids, clouds and web
services. These efforts resulted in tools among which
Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing}
and WS-PGRADE~\cite{Kacsuk2012}. Such workflow engines usually
describe applications in a high-level language with specific data and
control flow constructs, parallelization operators, visual edition
tools, links with domain-specific application repositories, provenance
recording and other features. An overview of workflow system
capabilities is available in~\cite{deelman2009workflows}.

@@

\begin{figure*}
\centering
\hspace*{0.2\columnwidth}
\subfloat[Tight integration]{
\def\svgwidth{0.6\columnwidth}
\input{figures/tight.pdf_tex}
\label{archi:tight}
}
\hfill
\subfloat[Service invocation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/service.pdf_tex}
\label{archi:service}
\hspace*{0.2\columnwidth}
}\\
\hspace*{0.2\columnwidth}
\subfloat[Task encapsulation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/sub-task.pdf_tex}
\label{archi:sub-task}
}
\hfill
\subfloat[Pool]{
\def\svgwidth{0.6\columnwidth}
\input{figures/agent.pdf_tex}
\label{archi:agent}
\hspace*{0.2\columnwidth}
}\\
\subfloat[Nested workflows. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-1.pdf_tex}
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-2.pdf_tex}
\label{archi:nested}
}\\
\subfloat[Workflow import. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.5\columnwidth}
\input{figures/import-1.pdf_tex}
\hfill
\def\svgwidth{0.5\columnwidth}
\input{figures/import-2.pdf_tex}
\label{archi:import}
}
\caption{Architectures}
\label{fig:architectures}
\end{figure*}
\begin{table*}
\centering
\footnotesize
\rowcolors{1}{white}{gray!25}
\begin{tabular}{ll}
  \textbf{Architecture} & \textbf{Systems} \\
  \hline
  Tight & \pbox{1.5\columnwidth}{
          Catania Science Gateway Framework~\cite{ardizzone2012decide},
          Distributed application runtime environment (DARE~\cite{maddineni2012distributed}),
          DECIDE~\cite{ardizzone2012decide}$^{c,n}$, LONI Pipeline Environment$^n$~\cite{dinov2009efficient}
          }
  \\
  Service & \pbox{1.5\columnwidth}{
            Apache Airvata~\cite{marru2011apache}, Neuroscience Gateway~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{kruger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
            Science gateways in~\cite{kacsuk2014science}$^w$
            } \\
  Task & CBRAIN and PSOM~\cite{GLAT-16}$^n$, CBRAIN and FSL$^n$\\
  Pool & SHIWA pool~\cite{ROGE-13}\\
  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$, HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517}, Tavaxy~\cite{Abouelhoda2012}.
           }\\
  Import & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$, Tavaxy~\cite{Abouelhoda2012}, system in~\cite{delaGarza2016}.
\end{tabular}
\caption{Classification of science gateways based on the architecture used to integrate workflow engines. $^c$: based on the Catania Science Gateway Framework. $^w$: based on WS-PGRADE/gUSE. $^n$: used for neuroimaging.}
\label{table:system-classification}
\end{table*}

@@

  the infrastructure, which may require administrative privileges;
  (\texttt{a$_2$}) the workflow is configured in the science gateway
  so that it becomes available to users. Note that integrating a
  workflow is not the same process as integrating a workflow
  \emph{engine}.

@@

See Figure~\subref*{archi:service}. The workflow engine is available
externally to the science gateway, as a service. The science gateway
controls the service through a specific interaction (\texttt{d}) that
might be implemented as a web-service call (e.g., RESTful or SOAP), as
a command-line or as any other method that offers a well-defined
interface to the workflow engine. The workflow engine might be invoked
either as a black box that completely masks the infrastructure and
tasks, or as a white box that allows for some interaction with
them. The workflow engine is responsible for controlling the tasks on
the infrastructure (\texttt{b}) and for performing the required data
transfers to execute them (\texttt{c$_2$}). User data is usually
managed through the science gateway (\texttt{c$_1$}), although in some
variations of the architecture (not represented in
Figure~\ref{fig:architectures}) it might as well be delivered by the
workflow engine directly to the user. \answerednote{Sil}{about c1: should it
  also show from engine to infra in the figure? this is what i
  understand from the above sentence}{Tristan}{Actually c1 is either
  coming from the science gatewray or from the workflow engine but
  then it merges with c2. I don't think we can cover both cases in a
  single architecture, the latter case would be another variation of
  this architecture. I modified a bit the text, is it clearer now?}

@@

See Figure~\subref*{archi:nested}. In nested workflows
(Figure~\subref*{archi:nested}-Left), workflows include other
workflows that are executed by a \emph{child} engine. Parent
and child engines might use different languages and might run on
different infrastructures. A parent workflow is
also called meta-workflow. The science gateway communicates with the
parent engine through abstract interaction \texttt{*$_1$}. The science
gateway also communicates with the infrastructure to transfer user
data through abstract interactions \texttt{*$_2$} and
\texttt{*$_3$}. Both workflow engines communicate with the
infrastructure through abstract interactions \texttt{*$_4$} and
\texttt{*$_5$}. The parent engine communicates with the child engine
through abstract interaction \texttt{*$_6$}. Administrator installs
workflows through interactions \texttt{a$_1$} and \texttt{a$_2$}.

@@

\subsection{Workflow import}

See Figure~\subref*{archi:import}. This is an abstract model
instantiated with the service invocation architecture for
consistency. Workflows are integrated in the science gateway through
format conversion from a native format to the science gateway
format. Workflow import is usually an offline process that is not
involved in workflow execution. A few systems have implemented
workflow import. In the SHIWA Simulation Platform, it is implemented
through the IWIR language, which provides a common language for
portability across grid workflow
systems~\cite{plankensteiner-prodan-etal:2013} and allows conversion
among $n$ workflow languages using $2n$ interactions instead of $n^2$.
Tavaxy~\cite{Abouelhoda2012}
enables the import, merging and execution of Taverna~\cite{oinn2004taverna} and
Galaxy~\cite{goecks2010galaxy} workflows; when some workflow parts
cannot be imported, workflows are run by Tavaxy as nested workflows
using their native engine. The work in~\cite{delaGarza2016} describes
workflow import from KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE and
from Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.

@@

                                     & \textbf{Import} \\

@@

  \textbf{Total} (missing features) & \cellcolor[HTML]{99CC99}\textbf{1}
                                     & \cellcolor[HTML]{99CC99}\textbf{1}

@@

                                     & \cellcolor[HTML]{99CC99}\textbf{1}
                                     & \cellcolor[HTML]{99CC99}\textbf{1}
                                     & \cellcolor[HTML]{99FF99}\textbf{0}\\

@@

\emph{Robustness of workflow execution} measures the likelihood that the
workflow execution fails due to errors in the components or in the
interactions of the software architecture. Errors coming
from the infrastructure (e.g., unavailable data or terminated tasks)
or workflows (e.g., wrong user input or application errors) are not
covered, since they do not stem directly from the software
architecture. Robustness is measured here as a consequence of global
complexity, since complex architectures tend to be more prone to
failure. More precisely, robustness is determined by the number of
software components and interactions represented in
red in the architecture diagrams in
Figure~\ref{fig:architectures}, which are specific to workflow
execution. Two metrics are used:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Specific components  (\texttt{R$_1$}): number of specific components
involved in workflow executions. The Science
gateway component, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc., and
\item Specific interactions (\texttt{R$_2$}): number of specific
  interactions involved in workflow executions. Data control
  (interaction \texttt{c}), for instance, is \emph{not} specific to
  workflow execution since it is also used by the science gateway to
  transfer user data. 
\end{itemize}
\texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the global score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution.

@@

\emph{Scalability} corresponds to the ability of an architecture to
cope with high workloads through the following mechanisms: starting multiple engine instances, distributing engines and task scheduling. 
This is measured by assessing the potential
scalability difficulties due to (partial) absence of a given feature in the architecture.  
Features are evaluated using a 3-level metric: 0 means that
the feature is very easy to enable, 1 means that it can be implemented
but with some difficulty, and 2 means that the feature cannot
 be implemented in a realistic scenario. Three different features are identified:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Multiple engine instances (\texttt{S$_1$}): measures the potential of having
  more than one engine instance in the architecture simultaneously. Workflow engines
  may require important amounts of resources when several workflows,
  or large workflows, are executed. At some point, it may be required
  for the science gateway to means that adding a new engine instance is a
  fully automated process, i.e. the workflow engine only has to be
  started. In this case, elastic engines can be implemented through some kind
  of auto-scaling mechanism to control the number
  of engine instances in the architecture. \texttt{S$_1$}=1 means that
  adding a new engine instance requires some form of manual
  intervention, which prevents easy
  implementations of elastic engines. \texttt{S$_1$}=2 means that new engine
  instances cannot be added.
\item Distributed engines (\texttt{S$_2$}): measures the potential of distributing
  the execution \emph{of a single workflow} among different engine
  instances. In our scope, this feature focuses on the capabilities of
  the architecture rather than these of the workflow
  engine. \texttt{S$_2$}=0 means that distributed engines are enabled by
  the architecture, and \texttt{S$_2$}=1 that they require specific
  developments in the workflow engine. 
\item Task scheduling (\texttt{S$_3$}): measures the complexity
  introduced by the architecture regarding task scheduling on
  the infrastructure. Task
  scheduling typically depends more on the implementation of specific
  algorithms in the science gateway, workflow engine and
  infrastructure than on the architecture used to integrate the
  workflow engine in the science gateway. Some architectures, however,
  complicate the task scheduling problem by introducing additional
  software layers or creating tasks with specific
  characteristics. \texttt{S$_3$}=0 means that the architecture does
  not introduce any additional complexity to the scheduling problem,
  and \texttt{S$_3$}=1 otherwise.
\end{itemize}
These three metrics are summed to obtain a global measure of the scalability
potential of the architecture.

\emph{Specific features} include:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Meta-workflow (\texttt{O$_1$}): measures the ability to describe
  meta-workflows from existing workflows. Meta-workflows offer an
  additional level of flexibility to build workflows from reusable
  components. \texttt{O$_1$}=0 means that meta-workflows are enabled by the architecture, and
  \texttt{O$_1$}=1 otherwise.
  \note{si}{why here we do not have the "possible to implement" situation?}

\item Fine-grained debugging (\texttt{O$_2$}): availability of
  fine-grained debugging information about workflow tasks (white-box
  workflow).  Fine-grained information about workflow tasks is
  required to properly troubleshoot workflow executions.
  \texttt{O$_2$}=0 means that such information is easy to access, and
  \texttt{O$_2$}=1 otherwise.
\end{itemize}
\texttt{O$_1$} and \texttt{O$_2$} are summed to obtain a total number of missing specific
features in the architecture.

\note{sil}{since we are dreaming, it seems to me that keeping provenance properly is also something to consider. the architecture can hide some details, and make it nearly impossible to keep provenance without hard coded hacks across the layers/components - i think}

>>>>

\author[mcgill]{Reza Adalat}
\author[mcgill]{Natacha Beck}
\author[mcgill]{\\Samir Das}
\author[isi]{Rafael Ferreira da Silva}
\author[mcgill]{Najmeh Khalili-Mahani}
\author[stpetersburg]{Vladimir Korkhov}
\author[criugm]{Pierre-Olivier Quirion}
\author[mcgill]{Pierre Rioux}
\author[amc]{S\'ilvia D. Olabarriaga}
\author[criugm]{Pierre Bellec}
\author[mcgill]{Alan C. Evans}

\address[mcgill]{McGill Centre for Integrative Neuroscience, Montreal Neurological Institute, McGill University, Canada.}
\address[creatis]{University of Lyon, CNRS, INSERM, CREATIS, Villeurbanne, France.}
\address[criugm]{Centre de Recherche de l'Institut de G\'eriatrie de Montr\'eal CRIUGM, Montr\'eal, QC, Canada.}
\address[isi]{University of Southern California, Information Sciences Institute, Marina del Rey, CA, USA.}
\address[amc]{Academic Medical Center of the University of Amsterdam, Department of Clinical Epidemiology, Biostatistics and Bioinformatics, Amsterdam, NL.}
\address[stpetersburg]{St. Petersburg State University, Russia.}

\begin{abstract}
  Workflow engines are critical for the efficient and transparent
  exploitation of distributed infrastructures in the ecosystem of
  tools and services offered by science gateways.  We investigate six
  software architectures commonly used to integrate workflow engines
  into science gateways. In \emph{tight integration}, the workflow
  engine shares software components with the science gateway. In
  \emph{service invocation}, the engine is isolated and invoked
  through a specific software interface. In \emph{task encapsulation},
  the engine is wrapped as a computing task executed on the
  infrastructure. In the \emph{pool model}, the engine is bundled in
  an agent that connects to a central pool to fetch and execute
  workflows. In \emph{nested workflows}, the engine is integrated as a
  child process of another engine. In \emph{workflow conversion}, the
  engine is integrated through workflow language conversion. We
  describe and evaluate these architectures with metrics for
  assessment of integration complexity, robustness, extensibility,
  scalability and support for meta-workflows and fine-grained
  debugging. Tight integration and task encapsulation are the easiest
  to integrate and the most robust. Extensibility is equivalent in
  most architectures. The pool model is the most scalable one and
  meta-workflows are only available in task encapsulation, nested
  workflows and workflow conversion. These results provide insights
  for science gateway architects and developers.
\end{abstract}

@@

This paper describes, provides examples and compares such
architectures, based on system-independent representations of their
main components and interactions. It is informed
\closednote{vladimir}{is this a good word for this context,
  'informed'? maybe 'based on'?}{Tristan}{I guess it is the right
  word, keeping it as is.} by our experience in the development and
sustained operation of the CBRAIN~\cite{SHER-14} and
VIP~\cite{GLAT-13} science gateways for medical image analysis during
the past 7 years, as well as by lessons learned from several science
gateway and workflow projects such as
SHIWA\footnote{\url{http://www.shiwa-workflow.eu}} and
ER-flow\footnote{\url{http://www.erflow.eu}}.  In this section we
provide background information and definitions of workflow engines,
science gateways and infrastructures. In
Section~\ref{sec:architectures}, we describe six architectures
\closednote{vladimir}{why is it exactly six? are these architectures
  the ones that happened to be implemented during various projects, or
  is it a full set of all possibilities (I believe not)? I think, it
  must be stated clearly why the focus is given to these particular
  ones - we are reviewing and evaluating the approaches that have been
  used already in our projects}{Tristan}{Fixed, thanks.} that cover
the main patterns that we encountered in the various systems that we
studied during the last few years. The architectures are described
within a consistent framework that underlines the functional
interactions between their main software components. In
Section~\ref{sec:evaluation}, the different architectures are
evaluated with metrics measuring integration complexity, robustness,
extensibility, scalability and other specific features. These metrics
are specifically designed to measure the ability of the architectures
to address workflow-related issues commonly encountered in science
gateways. \closednote{vladimir}{can we explain why exactly these
  metrics are used but not anything else? I think, we need to justify
  that these ones are especially important for the SG case; it would
  be nice to refer to some related research on software architecture
  metrics, Google gives about 2 million pages on this, where you can
  find metrics like cohesion, depth of scenario and so
  on}{Tristan}{The metrics are specifically designed to cover our
  particular workflow-SG context. Do you have any other
  suggestion?} We finally compare the architectures and highlight
recommendations for science gateway architects and 
developers.

@@

In the last decade, the e-Science community has developed workflow
systems to help application developers access distributed
infrastructures such as clusters, grids, clouds and web
services. These efforts resulted in tools among which
Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
WS-PGRADE~\cite{Kacsuk2012} and
WS-VLAM~\cite{wsvlam}. \closednote{vladimir}{As a past developer of it, I
  would suggest to refer to WS-VLAM~\cite{wsvlam} as well, despite it
  is not very famous, but it played an important role for Amsterdam
  e-science workflows}{Tristan}{Done, thanks.} Such workflow engines usually describe
applications in a high-level language with specific data and control
flow constructs, parallelization operators, visual edition tools,
links with domain-specific application repositories, provenance
recording and other features. An overview of workflow system
capabilities is available in~\cite{deelman2009workflows}.

@@

\begin{figure*}
\centering
\hspace*{0.2\columnwidth}
\subfloat[Tight integration]{
\def\svgwidth{0.6\columnwidth}
\input{figures/tight.pdf_tex}
\label{archi:tight}
}
\hfill
\subfloat[Service invocation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/service.pdf_tex}
\label{archi:service}
\hspace*{0.2\columnwidth}
}\\
\hspace*{0.2\columnwidth}
\subfloat[Task encapsulation]{
\def\svgwidth{0.6\columnwidth}
\input{figures/sub-task.pdf_tex}
\label{archi:sub-task}
}
\hfill
\subfloat[Pool]{
\def\svgwidth{0.6\columnwidth}
\input{figures/agent.pdf_tex}
\label{archi:agent}
\hspace*{0.2\columnwidth}
}\\
\subfloat[Nested workflows. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-1.pdf_tex}
\def\svgwidth{0.9\columnwidth}
\input{figures/nested-2.pdf_tex}
\label{archi:nested}
}\\
\subfloat[Workflow conversion. Left: abstract model. Right: instantiation with service invocation.]{
\def\svgwidth{0.5\columnwidth}
\input{figures/import-1.pdf_tex}
\hfill
\def\svgwidth{0.5\columnwidth}
\input{figures/import-2.pdf_tex}
\label{archi:import}
}
\caption{Architectures}
\label{fig:architectures}
\end{figure*}
\begin{table*}
\centering
\footnotesize
\rowcolors{1}{white}{gray!25}
\begin{tabular}{ll}
  \textbf{Architecture} & \textbf{Systems} \\
  \hline
  Tight & \pbox{1.5\columnwidth}{
          Catania Science Gateway Framework~\cite{ardizzone2012decide},
          Distributed application runtime environment (DARE~\cite{maddineni2012distributed}),
          DECIDE~\cite{ardizzone2012decide}$^{c,n}$, LONI Pipeline Environment$^n$~\cite{dinov2009efficient}
          }
  \\
  Service & \pbox{1.5\columnwidth}{
            Apache Airvata~\cite{marru2011apache}, Neuroscience Gateway~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{kruger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012},
            Science gateways in~\cite{kacsuk2014science}$^w$
            } \\
  Task & CBRAIN and PSOM~\cite{GLAT-16}$^n$, CBRAIN and FSL$^n$\\
  Pool & SHIWA pool~\cite{ROGE-13}\\
  Nested & \pbox{1.5\columnwidth}{SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$, HubZero with Pegasus (via hierarchical workflows)~\cite{Deelman201517}, Tavaxy~\cite{Abouelhoda2012}.
           }\\
  Conversion & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$, Tavaxy~\cite{Abouelhoda2012}, system in~\cite{delaGarza2016}.
\end{tabular}
\caption{Classification of science gateways based on the architecture used to integrate workflow engines. $^c$: based on the Catania Science Gateway Framework. $^w$: based on WS-PGRADE/gUSE. $^n$: used for neuroimaging.}
\label{table:system-classification}
\end{table*}

@@

  the infrastructure or prepared for on-the-fly installation, which
  may require specific privileges; \closednote{vladimir}{or bundled in
    packages considering all needed dependencies than can be deployed
    on the fly during the workflow execution}{Tristan}{added, thanks!} (\texttt{a$_2$}) the
  workflow is configured in the science gateway so that it becomes
  available to users. Note that integrating a workflow is not the same
  process as integrating a workflow \emph{engine}.

@@

See Figure~\subref*{archi:service}. The workflow engine is available
externally to the science gateway, as a service. The science gateway
controls the service through a specific interaction (\texttt{d}) that
might be implemented as a web-service call (e.g., RESTful or SOAP), as
a command-line or as any other method that offers a well-defined
interface to the workflow engine. The workflow engine might be invoked
either as a black box that completely masks the infrastructure and
tasks, or as a white box that allows for some interaction with
them. The workflow engine is responsible for controlling the tasks on
the infrastructure (\texttt{b}) and for performing the required data
transfers to execute them (\texttt{c$_2$}). User data is usually
managed through the science gateway (\texttt{c$_1$}), although in some
variations of the architecture (not represented in
Figure~\ref{fig:architectures}) it might as well be delivered by the
workflow engine directly to the user. \closednote{Sil}{about c1: should it
  also show from engine to infra in the figure? this is what i
  understand from the above sentence}{Tristan}{Actually c1 is either
  coming from the science gatewray or from the workflow engine but
  then it merges with c2. I don't think we can cover both cases in a
  single architecture, the latter case would be another variation of
  this architecture. I modified a bit the text to reflect this.}

@@

See Figure~\subref*{archi:nested}. In nested workflows
(Figure~\subref*{archi:nested}-Left), a workflow engine is integrated
as a \emph{child} process of a \emph{parent} workflow engine. Parent
and child engines might use different languages and might run on
different infrastructures. A parent workflow is also a
meta-workflow. The science gateway communicates with the parent engine
through abstract interaction \texttt{*$_1$}. The science gateway also
communicates with the infrastructure to transfer user data through
abstract interactions \texttt{*$_2$} and \texttt{*$_3$}. Both workflow
engines communicate with the infrastructure through abstract
interactions \texttt{*$_4$} and \texttt{*$_5$}. The parent engine
communicates with the child engine through abstract interaction
\texttt{*$_6$}. Administrator installs workflows through interactions
\texttt{a$_1$} and \texttt{a$_2$}.

@@

\subsection{Workflow conversion}

See Figure~\subref*{archi:import}. This is an abstract model
instantiated with the service invocation architecture for
consistency. Workflow engines are integrated in the science gateway
through workflow format conversion from a native format to the science
gateway format. \closednote{vladimir}{Are we talking only about importing
  workflows from other wf languages? I would suggest to modify the
  title of the section to 'Workflow conversion and import' then, as
  some gateways use the term 'import' to import their own native
  workflows, like WS-PGRADE, and using import only in the context of
  converting workflows can be confusing}{Tristan}{Good catch, the term
  'import' is actually a bit vague. I guess 'conversion' is more
  accurate. Replaced it everywhere in the text.} Workflow conversion
is usually an offline process that is not involved in workflow
execution. A few systems have implemented workflow conversion. In the
SHIWA Simulation Platform, it is implemented through the IWIR
language, which provides a common language for portability across grid
workflow systems~\cite{plankensteiner-prodan-etal:2013} and allows
conversion among $n$ workflow languages using $2n$ interactions
instead of $n^2$.  Tavaxy~\cite{Abouelhoda2012} enables the import,
merging and execution of Taverna~\cite{oinn2004taverna} and
Galaxy~\cite{goecks2010galaxy} workflows; when some workflow parts
cannot be imported, workflows are run by Tavaxy as nested workflows
using their native engine. The work in~\cite{delaGarza2016} describes
workflow conversion from KNIME~\cite{Berthold2008} to WS-PGRADE/gUSE
and from Galaxy~\cite{goecks2010galaxy} to WS-PGRADE/gUSE.

@@

                                     & \textbf{Conversion} \\

@@

  \textbf{Total} (missing features) & \cellcolor[HTML]{99FF99}\textbf{1}

@@

                                     & \cellcolor[HTML]{99FF99}\textbf{1}\\

@@

\emph{Robustness of workflow execution} measures the likelihood that the
workflow execution fails due to errors in the components or in the
interactions of the software architecture. Errors coming
from the infrastructure (e.g., unavailable data or terminated tasks)
or workflows (e.g., wrong user input or application errors) are not
covered, since they do not stem directly from the software
architecture. 
\closednote{vladimir}{I believe, proper error handling through the different levels of wf engine/gateway must be especially mentioned even if not especially covered. One of serious issues we faced while working with gateways at the AMC, in particular with gUse/WS-PGRADE, was that when it failed sometimes it was completely not clear what was the problem exactly. So we might consider discussing not only the number of software components but how explicit and clear interactions are between them; how transparently we can monitor the execution through all the levels and thus detect and prevent possible malfunction}{Tristan}{In our current evaluation framework, this is supposed to be covered by O2. I still agree with you that O2 could be further refined. See Github issue #27}
Robustness is measured here as a consequence of global
complexity, since complex architectures tend to be more prone to
failure. More precisely, robustness is determined by the number of
software components and interactions represented in
red in the architecture diagrams in
Figure~\ref{fig:architectures}, which are specific to workflow
execution. Two metrics are used:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Specific components  (\texttt{R$_1$}): number of specific components
involved in workflow executions. The Science
gateway component, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc., and
\item Specific interactions (\texttt{R$_2$}): number of specific
  interactions involved in workflow executions. Data control
  (interaction \texttt{c}), for instance, is \emph{not} specific to
  workflow execution since it is also used by the science gateway to
  transfer user data. 
\end{itemize}
\texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the global score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution.

@@

\emph{Scalability} corresponds to the ability of an architecture to
cope with high workloads through the following mechanisms: starting multiple engine instances, distributing engines and task scheduling. 
This is measured by assessing the potential
scalability difficulties due to (partial) absence of a given feature in the architecture.  
Features are evaluated using a 3-level metric: 0 means that
the feature is very easy to enable, 1 means that it can be implemented
but with some difficulty, and 2 means that the feature cannot
 be implemented in a realistic scenario. Three different features are identified:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Multiple engine instances (\texttt{S$_1$}): measures the
  potential for a single gateway instance to use more than one engine
  instance simultaneously. Workflow engines may require important
  amounts of resources when several workflows, or large workflows, are
  executed. At some point, it may be required for the science gateway
  to distribute the load among several engines. \texttt{S$_1$}=0 means
  \closednote{vladimir}{is this the right word here, 'to
    means'?}{Tristan}{This sentence had been truncated, thanks.} that
  adding a new engine instance is inherently supported in the
  architecture. In this case, elastic engines can be implemented
  through some kind of auto-scaling mechanism to control the number of
  engine instances in the architecture. \texttt{S$_1$}=1 means that
  multiple engine instances requires specific developments in the
  science gateway or workflow engine. \texttt{S$_1$}=2 means that new
  engine instances cannot be added.
\item Distributed engines (\texttt{S$_2$}): measures the potential of distributing
  the execution \emph{of a single workflow} among different engine
  instances. In our scope, this feature focuses on the capabilities of
  the architecture rather than these of the workflow
  engine. \texttt{S$_2$}=0 means that distributed engines are enabled by
  the architecture, and \texttt{S$_2$}=1 that they require specific
  developments in the workflow engine. 
\item Task scheduling (\texttt{S$_3$}): measures the complexity
  added by the architecture to task scheduling on
  the infrastructure. Task
  scheduling typically depends more on the implementation of specific
  algorithms in the science gateway, workflow engine and
  infrastructure than on the architecture used to integrate the
  workflow engine in the science gateway. Some architectures, however,
  complicate the task scheduling problem by introducing additional
  software layers or creating tasks with specific
  characteristics. \texttt{S$_3$}=0 means that the architecture does
  not add any additional complexity to the scheduling problem,
  and \texttt{S$_3$}=1 otherwise.
\end{itemize}
These three metrics are summed to obtain a global measure of the scalability
potential of the architecture.

\emph{Specific features} include:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Meta-workflow (\texttt{O$_1$}): measures the ability to describe
  meta-workflows from existing workflows. Meta-workflows offer an
  additional level of flexibility to build workflows from reusable
  components. \texttt{O$_1$}=0 means that meta-workflows are
  intrinsically enabled by the architecture (i.e. they can be
  implemented using the components and interactions already in place),
  amd \texttt{O$_1$}=1 means that they may be implemented with some
  development effort. \closednote{si}{why here we do not have the
    "possible to implement" situation?}{Tristan}{O$_1$=2 would never
    be reached. Besides, that would make it difficult to sum O1 and O2
    because they wouldn't measure the same type of thing. In the
    current situation, we are only talking about the presence/absence
    of a particular feature and this is why O1 and O2 can be added to
    obtain a total number of 'missing features'.}
  \closednote{vladimir}{I agree with Silvia, we should have the
    intermediate values as well, since meta-workflows might be
    straightforwardly supported by the gateway, or they are still
    possible to implement but only with some workarounds. Like it was
    with SHIWA: first 6 months coarsed-grained interoperability looked
    more like a workaround till it was better integrated and supported
    by WS-PGRADE itself}{Tristan}{Yes but in the situation you
    describe, evolutions came from the wf engine implementation rather
    than from the architecture to integrate it, which is our point
    here.}

\item Fine-grained debugging (\texttt{O$_2$}): availability of
  fine-grained debugging information about workflow tasks (white-box
  workflow).  Fine-grained information about workflow tasks is
  required to properly troubleshoot workflow executions. Its
  availability and accuracy usually depends on the number of software
  layers between the science gateway and the workflow
  engine~\cite{olabarriaga2014}. \texttt{O$_2$}=0 means that
  the information is directly accessible in the science gateway,
  \texttt{O$_2$}=1 means that it is obtained through one or more software
  interactions. \closednote{vladimir}{otherwise - does it mean the
    fine-grained info is available but hard to access or not available
    it all? the latter happens as well}{Tristan}{Precised the text.}
\end{itemize}
\texttt{O$_1$} and \texttt{O$_2$} are summed to obtain a total number of missing specific
features in the architecture.
